---
title: "Parse all annotation data into table"
author: Katelyn Nguyen (+Sections from Rachel Smoak, Bin He)
output:
    pdf_document: default
params:
    args: ''
---

```{=html}
<!--
To run this R Markdown from the command line with arguments:

Rscript -e "rmarkdown::render('parse_all.Rmd', params = list(args = '<output_directory> <01_output_file>'))"
-->
```

Some sections adapted from Rachel Smoak, Bin He's work in C037-Cand-auris-adhesin/01-global-adhesin-prediction /main-analysis.Rmd

## Parse hmmscan --domtblout output

```{r read_hmmscan}
library(dplyr)
library(tidyr)
library(ggplot2)
library(readr)

params <- list(args = "../results/aurisComplete ../../01-three-part-adhesin-test/results/proteinTable.csv")

args <- strsplit(params$args, " ")[[1]]

if (length(args) != 2){
  print("Usage: <input_directory> <01_output_file>")
  stop()
}

hmmscan_domtblout_file = paste0(args[1], "/hmmscan_domtblout")
three_part_test_file = args[2]
freak_output_directory = paste0(args[1], "/freak-output")
ser_thr_output = paste0(args[1], "/ST_freq_perProtein")
tango_output_directory = paste0(args[1], "/tango-output")
xstream_output_directory = paste0(args[1], "/tandem-repeats")

hmm.names <- c("target_name", "target_accession", "target_length", "query_name", "query_accession", "query_length", "seq_evalue", "seq_score", "seq_bias", "domain_num", "num_target_domains", "c_evalue", "i_evalue", "domain_score", "domain_bias", "hmm_from", "hmm_to", "ali_from", "ali_to", "env_from", "env_to", "accuracy")
hmmscan_df <- read_table(file = hmmscan_domtblout_file, col_names = hmm.names, col_types = "cciccidddiiddddiiiiiid", skip = 3)

hmmscan_df <- hmmscan_df %>%
  # Extract protein name
  separate(query_name, into = c(NA, "ID", "Name"), sep = "\\|")
```

### Filter out anything that doesn't satisfy inclusion threshold (seq_evalue \< 0.01 and c-evalue \< 0.01), Create pfam_domains, unique_domains, and uncertain_domains

```{r pfam_domains}
filtered_hmm_df <- hmmscan_df %>% 
  filter(seq_evalue < 0.01 & c_evalue < 0.01) %>% 
  group_by(ID) %>%
  # Collapse into list of pfam domains for each protein
  reframe(pfam_domains = paste0(target_name, " (", env_from, "-", env_to,  ")", collapse = ", "), 
          unique_domains = paste(unique(target_name), collapse = ", "))

# Filter uncertain domains
uncertain_domains_df <- hmmscan_df %>%
  filter(seq_evalue >= 0.01 | c_evalue >= 0.01) %>%
  distinct(ID, target_name, seq_evalue)

uncertain_domains_df <- uncertain_domains_df %>% 
  left_join(filtered_hmm_df %>% 
              select(ID, unique_domains), by = "ID") %>%
  rowwise() %>%
  # Remove uncertain domains already in unique_domains
  filter(!target_name %in% strsplit(unique_domains, ", ")[[1]]) %>%
  ungroup() %>%
  arrange(ID, seq_evalue) %>%   # Sort by ID then E-value (ascending)
  group_by(ID) %>%
  reframe(`uncertain_domains (E-value)` = paste0(target_name, " (", seq_evalue, ")", collapse = ", "))

# Merge back into main dataframe
filtered_hmm_df <- filtered_hmm_df %>%
  full_join(uncertain_domains_df, by = "ID") %>%
  relocate(`uncertain_domains (E-value)`, .after = last_col())
```

## Parse output from 01-three-part-adhesin-test

```{r}
part_one_df <- read_csv(file = three_part_test_file)
part_one_df <- part_one_df %>% rename(ID = `Protein ID`)

# Sort
part_one_df <- part_one_df[order(-part_one_df$"FungalRV Score",
                                 -part_one_df$"Signal Peptide",
                                 -as.numeric(part_one_df$"GPI-anchor")), ]

# Add failed_tests column to signify which of the 3 tests failed
part_one_df <- part_one_df %>%
  mutate(
    failed_tests = paste(
      ifelse(`FungalRV Score` <= 0.511, "FungalRV", ""),
      ifelse(`Signal Peptide` <= 0.5, "SignalP", ""),
      ifelse(`GPI-anchor` != TRUE, "PredGPI", ""),
      sep = ", ")
  ) %>%
  mutate(
    failed_tests = gsub(", ,", ",", failed_tests),
    failed_tests = gsub("^, |, , |, $", "", failed_tests)
  )
```

## Combine hmmscan, 01-three-part-test output

```{r}
# Merge hmmscan table and 01_output table
all_hmm_df <- data.frame()
all_hmm_df <- merge(filtered_hmm_df, part_one_df[, c("ID", "FungalRV Score", "failed_tests")], by = "ID", all.y = TRUE)
# Sort by FungalRV score
all_hmm_df <- all_hmm_df[order(-all_hmm_df$"FungalRV Score"), ]
```

## Read Ser/Thr frequencies

```{r}
# (1) Panproteome-wide: Maximum frequency of Ser/Thr, Ser, Thr in 100aa window
ST.freq <- read_tsv(paste0(freak_output_directory,"/ST_freq_freak.out"), col_types = "cid")
S.freq <- read_tsv(paste0(freak_output_directory,"/S_freq_freak.out"), col_types = "cid")
T.freq <- read_tsv(paste0(freak_output_directory,"/T_freq_freak.out"), col_types = "cid")
ST.window <- bind_rows("ST" = ST.freq, "S" = S.freq, "T" = T.freq, .id = "residue") %>% 
  group_by(id, residue) %>% 
  summarize(max = max(freq))
rm(list = c("ST.freq", "S.freq", "T.freq"))

ST.window %>% 
  left_join(select(part_one_df, id = "Protein Name", failed_tests), by = "id") %>% 
  mutate(residue = factor(residue, levels = c("ST", "S", "T"), labels = c("Ser/Thr", "Ser", "Thr")),
         `Passed Tests` = ifelse(failed_tests == "", "Yes", "No")) %>% 
  ggplot(aes(x = max)) + geom_histogram(binwidth = 0.02) + 
  xlab("Max frequency of Ser/Thr in 100 amino acid window") + scale_y_continuous(position = "right") +
  facet_grid(`Passed Tests` ~ residue, scales = "free_y", labeller = "label_both", switch = "y") + theme_minimal(base_size = 12) +
  theme(
    strip.placement = "outside",
    strip.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    panel.spacing = unit(1, "lines")
  )

# (2) Panproteome-wide: Ser, Thr frequencies over whole protein
ST.protein <- read_tsv(paste0(ser_thr_output, col_types = cols()))
# Extract protein ID, protein name
ST.protein <- ST.protein %>% 
  separate(ID, into = c("tmp", "ID", "Name"), sep = "\\|") %>%
  select(-tmp)

ST.protein %>% 
  left_join(select(part_one_df, `ID`, failed_tests), by = "ID") %>% 
  mutate(`Ser/Thr` = (Ser+Thr)/length, Ser = Ser/length, Thr = Thr/length) %>% 
  pivot_longer(cols = c(`Ser/Thr`, Ser, Thr), names_to = "residue", values_to = "value") %>% 
  mutate(residue = factor(residue, levels = c("Ser/Thr", "Ser", "Thr")),
         `Passed Tests` = ifelse(failed_tests == "", "Yes", "No")) %>% 
  ggplot(aes(x = value)) + geom_histogram(binwidth = 0.02) + 
  xlab("Frequency of Ser/Thr over each protein") + scale_y_continuous(position = "right") +
  facet_grid(`Passed Tests` ~ residue, scales = "free_y", labeller = "label_both", switch = "y") + theme_minimal(base_size = 12) +
  theme(
    strip.placement = "outside",
    strip.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    panel.spacing = unit(1, "lines")
  )

# (3) Merge with all_hmm_df
# Combine Maximum and Protein-wide Frequencies
tmp <- ST.protein %>%
  mutate(ST.prot = round((Ser+Thr)/length, 3)) %>%
  select(ID, Name, ST.prot) %>%
  left_join(ST.window %>% filter(residue == "ST") %>%
  select(Name = id, ST.window.max = max), by = "Name") %>%
  # for proteins shorter than 100 a.a., the sliding window estimate would be NA. replace it with whole protein estimate in ST.protein
  mutate(ST.window.max = round(coalesce(ST.window.max, ST.prot),3)) %>%
  rename("S/T Frequency" = ST.prot, "Max window S/T frequency" = ST.window.max)

# Add Maximum, Protein-wide Frequencies to all_hmm_df
all_hmm_df <- all_hmm_df %>%
  left_join(tmp, by = "ID") %>%
  relocate(Name, .after = 1)
```

## Read TANGO output
Average B-Aggregation per protein
```{r}
# Read in Average Aggregation per protein
tango_df <- read_tsv(paste0(tango_output_directory,"/perSequence_aggregation.txt")
                     , col_types = cols())
# Extract protein ID, protein name
tango_df <- tango_df %>% 
  separate(Sequence, into = c("tmp", "ID", "Name"), sep = "\\|") %>%
  select(-tmp)

# Merge with all_hmm_df
all_hmm_df <- all_hmm_df %>% 
  left_join(tango_df %>% 
              select(ID, "Average B-aggregation per residue" =  Aggregation)
            , by = "ID")
```

Using Bin's tango extraction functions

```{r extract_tango_info}
extract_tango <- function(tango_output, agg_threshold = 5, required_in_serial = 5) {
    require(tidyverse)
    tmp <- read_tsv(file = tango_output, col_types = "icddddd") %>% 
        # a boolean vector for residues above threshold
        mutate(pass = Aggregation > agg_threshold)
    pass.rle <- rle(tmp$pass) # this creates a run length encoding that will be useful for identifying the sub-sequences in a run longer than certain length
    # --- Explanation ---
    # this rle object is at the core of this function
    # an example of the rle looks like
    #   lengths: int[1:10] 5 19 20 8 1 5 19 6 181 18
    #   values: logi[1:10] F T  F  T F T F  T F   T
    #   note that by definition the values will always be T/F interdigited
    # our goal is to identify the sub-sequences that is defined as a stretch of 
    # n consecutive positions with a score greater than the cutoff and record the
    # sub-sequence, its length, start and end position, 90% quantile of the score
    # --- End of explanation ---
    # 1. assigns a unique id for each run of events
    tmp$group <- rep(1:length(pass.rle$lengths), times = pass.rle$lengths)
    # # 2. extract the subsequences
    agg.seq <- tmp %>%
        dplyr::filter(pass) %>% # drop residues not predicted to have aggregation potential
        group_by(group) %>% # cluster by the runs
        summarise(seq = paste0(aa, collapse = ""),
                  start = min(res), end = max(res), length = n(),
                  median = median(Aggregation),
                  q90 = quantile(Aggregation, probs = 0.9),
                  ivt = sum(aa %in% c("I","V","T")) / length(aa),
                  .groups = "drop") %>%
        mutate(interval = start - dplyr::lag(end) - 1) %>%
        dplyr::filter(length >= required_in_serial) %>%
        select(-group)
    return(agg.seq)
}
```

Apply to individual proteome

```{r tango_extract, eval=FALSE, warning=FALSE}
tango.output.files <- list.files(path = tango_output_directory, pattern = "^tr\\|.*\\.txt(\\.gz)?$", full.names = TRUE)
# the read_csv() function used in the custom function can automatically decompress gzipped files
tango.res <- lapply(tango.output.files, extract_tango)
names(tango.res) <- gsub(".txt|.txt.gz", "", basename(tango.output.files))
# to add species information
# seqInfo <- read_tsv("raw-output/XP_028889033_homologs.tsv", comment = "#", col_types = c("ccci"))
tango.res.df <- bind_rows(tango.res, .id = "id")
# Extract protein ID, protein name
tango.res.df <- tango.res.df %>% 
  separate(id, into = c("tmp", "ID", "Name"), sep = "\\|") %>%
  select(-tmp)
```

Add B-aggregation sequences (and medScore and IVT) to all_hmm_df
```{r merge_tango}
# ivt_30th <- quantile(tango.res.df$ivt, 0.3)
# median_30th <- quantile(tango.res.df$median, 0.3)

tmp <- tango.res.df %>%
  select(ID, seq, median, ivt) %>%
  group_by(ID) %>%
  # Filter out sequences with low median (< 30th percentile) and low IVT (< 30th percentile) scores
  # filter(ivt >= ivt_30th | median >= median_30th) %>%
  summarize("B-agg seq count" = n(), "B-agg sequences" = paste(sort(seq), collapse = ", "), 
            medScore = round(mean(median),1),
            IVT = round(mean(ivt),2),)

all_hmm_df <- all_hmm_df %>% 
  left_join(tmp %>% 
              select(ID,  "B-agg sequences", "B-agg seq count")
            , by = "ID") %>% 
  mutate(`B-agg seq count` = replace_na(`B-agg seq count`, 0))
```

```{r graph_beta_agg}
# Graph beta-aggregation seq count / aggregation per residue
all_hmm_df %>%
  ggplot(aes(x = (failed_tests == ""), y = `B-agg seq count`)) + 
  geom_violin() + 
  scale_x_discrete(labels = c("Failed", "Passed All Tests")) +
  theme_minimal(base_size = 12) +
  theme(
    strip.placement = "outside",
    strip.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    panel.spacing = unit(1, "lines")
  )

all_hmm_df %>%
  ggplot(aes(x = (failed_tests == ""), y = `Average B-aggregation per residue`)) + geom_violin() + 
  scale_x_discrete(labels = c("Failed", "Passed All Tests")) +
  theme_minimal(base_size = 12) +
  theme(
    strip.placement = "outside",
    strip.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    panel.spacing = unit(1, "lines")
  )
```

Top 20 TANGO hits ranked by number of occurrences in all proteins, restricted to the strong hits with a median score \>= 30%.
```{r unique_motifs}
# tango.res.df1 <- tango.res.df %>% left_join(all_hmm_df %>% select(ID), by = "ID")

# find unique motifs and count the number of proteins and strains represented
motif.summary <- tango.res.df %>% 
  # left_join(select(all_hmm_df, id = ID, isHIL)) %>% 
  left_join(select(all_hmm_df, ID)) %>% 
  group_by(seq) %>% 
  summarize(n = n(),
            n.prot = n_distinct(ID), 
            #n.strains = n_distinct(strain),
            medScore = round(mean(median),1),
            IVT = round(mean(ivt),2),
            avg.intv = round(mean(interval, na.rm = T),1), 
            mad.intv = round(mad(interval, na.rm = T),1),
            .groups = "drop") %>% 
  arrange(desc(n))
motif.summary %>% 
  filter(round(medScore,0) >= 30) %>% 
  head(20)
```

## Read XSTREAM output

```{r}
repeats_df <- read_tsv(paste0(xstream_output_directory, "/XSTREAM_proteins_i0.7_g3_m5_L15_chart.xls"), col_types = cols())
# Extract protein ID, protein name
repeats_df <- repeats_df %>% 
  separate(identifier, into = c("tmp", "ID", "Name"), sep = "\\|") %>%
  select(-tmp)

# Calculate per sequence repeat percentage and add to all_hmm_df
tmp <- repeats_df %>% 
  mutate(len = end - start + 1) %>% 
  group_by(ID) %>% 
  summarize(perc = round(sum(len/`seq length`),3)) %>% 
  rename("TR Percentage" = perc)

all_hmm_df <- all_hmm_df %>% 
  left_join(tmp, by = "ID") %>% 
  replace_na(list("TR Percentage" = 0))

# Graph TR Percentage
all_hmm_df %>%
  ggplot(aes(x = (failed_tests == ""), y = `TR Percentage`)) + geom_violin() + 
  scale_x_discrete(labels = c("Failed", "Passed All Tests")) +
  theme_minimal(base_size = 12) +
  theme(
    strip.placement = "outside",
    strip.background = element_blank(),
    panel.border = element_rect(color = "black", fill = NA),
    panel.spacing = unit(1, "lines")
  )
```
Parse and Save final tables
```{r}
all_hmm_df <- all_hmm_df %>% 
  relocate(`FungalRV Score`, .after = `Name`) %>%
  relocate(`failed_tests`, .after = `FungalRV Score`)

# Summary Table of Percentiles
percentages <- c(0.25, 0.50, 0.75, 0.90, 0.95, 0.99)
st_freq_col <- quantile(all_hmm_df$`S/T Frequency`, probs = percentages)
max_window_st_col <- quantile(all_hmm_df$`Max window S/T frequency`, probs = percentages)
b_agg_per_residue_col <- quantile(all_hmm_df$`Average B-aggregation per residue`, probs = percentages)
b_agg_count_col <- quantile(all_hmm_df$`B-agg seq count`, probs = percentages)
tr_col <- quantile(all_hmm_df$`TR Percentage`, probs = percentages)

percentile_df <- data.frame("Percentile" = percentages, "S/T Frequency" = st_freq_col, "Max Window S/T Frequency" = max_window_st_col, "Average B-aggregation Per Residue" = b_agg_per_residue_col, "B-aggregation seq count" = b_agg_count_col, "TR Percentage" = tr_col)

# Add row for mean
percentile_df <- percentile_df %>%
  rbind("Mean" = c(NA, mean(all_hmm_df$`S/T Frequency`), 
        mean(all_hmm_df$`Max window S/T frequency`), 
        mean(all_hmm_df$`Average B-aggregation per residue`), 
        mean(all_hmm_df$`B-agg seq count`), 
        mean(all_hmm_df$`TR Percentage`)))
```
Split into 3 Tables: Three (passing three conditions), Two (passing two conditions), and One (passing at least one condition)
```{r}
# Table 1: All 3 criteria satisfied (FungalRV, PredGPI, SignalP)
df_three <- all_hmm_df[all_hmm_df$failed_tests == "", ] %>%
  select(-failed_tests)
rownames(df_all) <- NULL

# Table 2: Two criteria satisfied (1 test failed)
df_two <- all_hmm_df[sapply(strsplit(all_hmm_df$failed_tests, split=","), length) == 1, ]
rownames(df_two) <- NULL

# Table 3: One criteria satisfied (2 tests failed)
df_one <- all_hmm_df[sapply(strsplit(all_hmm_df$failed_tests, split=","), length) == 2, ]
rownames(df_one) <- NULL

write.csv(df_three, file = paste0(args[1], "/df_three.csv"))
print("Annotated proteins table saved to df_three.csv")
write.csv(df_two, file = paste0(args[1],"/df_two.csv"))
print("Annotated proteins table saved to df_two.csv")
write.csv(df_one, file = paste0(args[1],"/df_one.csv"))
print("Annotated proteins table saved to df_one.csv")
```